---
title: "ggplot2 tutorial - R Bootcamp 2018"
author: "Alicia Vallorani"
date: "August 13, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Install packman for subsequent package management
```{r, include=FALSE}
if (!("pacman" %in% installed.packages()[,])) {
    install.packages("pacman")
  }
pacman::p_load(base, dplyr, tidyr, ggplot2, Hmisc, GGally, jtools)
```


## Set working directory and read-in data
```{r}
setwd("~/Penn State/Summer 2018/R Bootcamp 2018")

read.csv("./DATA/ggplot2_tutorial.csv", stringsAsFactors = FALSE) %>%
  mutate(bi = as.factor(bi),
         sex = as.factor(sex),
         socbid_group = as.factor(socbid_group),
         prosoc_group = as.factor(prosoc_group)) -> df
```


## Examine data structure
```{r}
str(df)
```


## Examining distributions - histograms
```{r}
# Looking at a histogram for a single variable
ggplot(df, aes(proportion_toys)) +
  geom_histogram(bins = 5) # you can change the bin value to best fit your data

# Looking at histrograms for all variables ggplot option
ggplot(gather(df[c(-1, -2)]), aes(value)) + # numbers are removing id columns
    geom_histogram(bins = 5) + 
    facet_wrap(~key, scales = "free_x")

# Looking at histrograms for all variables Hmisc option
hist.data.frame(df[c(-1, -2)]) 
```


## Examining zero-order relations - scatterplots
```{r}
# Looking at a scatterplot for a single relation
ggplot(df, aes(proportion_toys, proportion_socconv)) +
  geom_point()
  #stat_smooth() 
  #stat_smooth(method = "lm", color = "black", se = TRUE)
  #facet_wrap(~bi)
  #facet_wrap(~socbid_group)
  #face_grid(bi~sex)

# Looking at a scatterplot matrix
# Michael, I've never really found a ggplot option I am hapy with. Do you have good code?

# Fancy distribution/relations option with GGally
na.omit(df) -> df_na # removing na to reduce error messages
ggpairs(df_na[c(-1, -2)])
```


## Examining group differences - bar graphs
```{r}
ggplot(df, aes(bi, proportion_peerbody)) +
  geom_bar(stat = "identity") 
  #theme_bw() 
  #theme(panel.grid.major.x = element_blank(),
        #panel.grid.minor.x = element_blank(),
        #panel.grid.major.y = element_blank(),
        #panel.grid.minor.y = element_blank(),
        #panel.spacing = unit(1.5, "lines")) 
  #labs(x = "Behavioral Inhibition Group",
       #y = "Proportion of Time Looking at Peer Body") 
  #scale_x_discrete(limits = c("1", "0"),
                   #labels = c("BI", "BN"))
```


## Examining interactions - bar graphs
```{r}
ggplot(df, aes(bi, proportion_socconv, fill = sex)) +
  geom_bar(stat = "identity", position = "dodge")
  #labs(fill = "Sex") 
  #scale_fill_manual(labels = c("Boys", "Girls"), values = c("#ff8c00", "#5898d7")) 
  #theme(legend.justification=c(-0.1,1), legend.position=c(.75,.95)) 
  #coord_flip()
```


## Examining interactions - 3 variables and continuous interactions
```{r}
ggplot(df, aes(proportion_toys, proportion_socconv, color = sex)) +
  geom_point() +
  stat_smooth(method = lm, se = FALSE) 
  #facet_wrap(~bi)


# Advanced options using jtools
with(df, lm(proportion_socconv~biq*proportion_toys)) -> lm1

interact_plot(lm1, pred = biq, modx = proportion_toys,
              plot.points = TRUE,
              x.label = "Behavioral Inhibition",
              y.label = "Proportion of Time Engaging in Social Conversation",
              legend.main = "Prop Dwell Toys")

johnson_neyman(lm1, pred = biq, modx = proportion_toys, alpha = 0.05)
```


## Publication quality graphs and exporting 
```{r}
df %>%
  dplyr::select(id, bi, proportion_peerbody, proportion_socconv) %>%
  gather(key = type, value = proportion, proportion_peerbody, proportion_socconv)  -> social

  ggplot(social, aes(type, proportion, fill = bi)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_x_discrete(limits = c("proportion_peerbody", 
                              "proportion_socconv"), 
                   labels = c("Dwell Peer", "Social Conversation")) +
  labs(x = "Type of Behavior",
       y = "Proportion of Time Engaged in Behavior",
       fill = "Temperament") +
  scale_fill_manual(labels = c("BN", "BI"), values = c("#ff8c00", "#5898d7")) +
  theme_bw() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.spacing = unit(1.5, "lines")) +
  theme(axis.text = element_text(size = 12)) +
  theme(axis.title = element_text(size = 14)) +
  theme(legend.text = element_text(size = 12)) +
  theme(legend.title = element_text(size = 14)) +
  theme(legend.justification=c(-0.1,1), legend.position=c(.75,.95)) -> social_plot

ggsave("./OUTPUT/FIGURES/social_plot.png", plot = social_plot, 
       width = NA, height = NA, dpi = 300)

# For editable figures  
svg("./OUTPUT/FIGURES/social_plot.svg")
plot(social_plot)
dev.off()
```

