---
title: "ggplot2 tutorial - R Bootcamp 2018"
author: "Alicia Vallorani"
date: "August 13, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

#handy package manager that installs and loads packages used in this document
if (!("pacman" %in% installed.packages()[,"Package"])) { install.packages("pacman") } 
pacman::p_load(dplyr, tidyr, ggplot2, GGally, Rmisc, jtools) #install (if needed) and load packages
```


## Goals of this markdown
This markdown is designed to provide an introduction to data visualization in R. Primarily it will cover ggplot2; although a few advanced options are also covered. Questions about code can be directed to Alicia Vallorani (auv27@psu.edu).


## Read in data
```{r}
df <- read.csv("../data/ggplot2_tutorial_vallorani.csv", stringsAsFactors = FALSE) %>%
  mutate_at(vars(bi, sex, socbid_group, prosoc_group), as.factor) 
```


## Examine data structure
These data are drawn from a cross-sectional study assessing attention to the social environment and socio-emotional behaviors. Children are 5-7 and complete a social dyad with a novel peer. Children are of differing temperament: one fearful (BI) and one non-fearful (BN). Mobile eyetracking and behavioral data are collected across the 5 free-play interaction. General expecations for analysis would be to see BI children engaging in fewer social behaviors than BN children.

### Describe data structure
id: participant id
did: dyad id
sex: 1 = boy; 2 = girl
bi: 0 = BN; 1 = BI
biq: continuous measure of temperamental fearfulness
proportion_peerbody: dwell time to peer body
proportion_peerface: dwell time to peer face
proportion_self: dwell time to self
proportion_toys: dwell time to toys
proportion_other: dwell time to other stimuli
proportion_socconv: time in social conversation (discussion about topics other than play)
proportion_playconv: time in play conversation (discussion about ongoing play)
socialbid_group: 0 = few social bids; 1 = medium social bids; 2 = many social bids
prosoc_group: 0 = few prosocial behaviors; 1 = medium prosocial behaviors; 2 = many prosocial behaviors
```{r}
str(df)
```


## Examining univariate distributions - histograms
This section walks through making histograms for single and multiple variables.
```{r}
# Looking at a histogram for a single variable
ggplot(df, aes(proportion_toys)) +
  geom_histogram(bins = 5) # you can change the bin value to best fit your data

# Looking at histograms for all variables ggplot option
ggplot(df %>% select(starts_with("proportion"), age) %>% # selecting non-binary variables
         gather(), aes(value)) + # grouping for visualization
    geom_histogram(bins = 5) + 
    facet_wrap(~key, scales = "free_x") # free_x allows for differing x-axes 
```


## Examining zero-order relations - scatterplots
This section walks through how to make a simple scatterplot between two variables. Additionally, you can add a fit line and look at how scatterplots may vary across groups.
```{r, fig.height=5, fig.width=10, messages=FALSE, warnings=FALSE}
ggplot(df, aes(x=proportion_toys, y=proportion_socconv)) +
  geom_point()
  #stat_smooth() 
  #stat_smooth(method = "lm", color = "black", se = TRUE) 
  #facet_wrap(~bi) 
  #facet_wrap(~socbid_group)
  #face_grid(bi~sex)
```


## Examining zero-order relations - scatterplot matrices 
This section walks through examining multiple relations simultaneoulsly. These plots are created using GGally which couples with ggplot2.
```{r, fig.height=4, fig.width=7, messages=FALSE, warnings=FALSE}
ggpairs(df %>%
          select(starts_with("proportion")) %>%
                 na.omit(), progress=FALSE, lower = list(combo = wrap("facethist", bins=6)))

# Matrix including dichotomous variables
ggpairs(df %>% 
          dplyr::select(bi, socbid_group, proportion_peerbody, 
                 proportion_socconv) %>% # subsetting for visualization
          dplyr::rename(socbid = socbid_group, peerbody = proportion_peerbody, 
                 socconv = proportion_socconv) %>% # renaming variables for figure
          na.omit(), progress=FALSE, lower = list(combo = wrap("facethist", bins=6)))
        #mapping = aes(color = bi)) # coloring the plot by group
```


## Examining group differences - bar graphs
This section walks through a basic bar graph. Once the base graph is made, you can make edits to the theme and axes to your preference. You can also include error bars
```{r}
ggplot(df, aes(x = bi, y = proportion_peerbody)) +
  geom_bar(stat = "identity") 
  #theme_bw() 
  #theme(panel.grid.major.x = element_blank(),
        #panel.grid.minor.x = element_blank(),
        #panel.grid.major.y = element_blank(),
        #panel.grid.minor.y = element_blank(),
        #panel.spacing = unit(1.5, "lines")) 
  #labs(x = "Behavioral Inhibition Group",
       #y = "Proportion of Time Looking at Peer Body") 
  #scale_x_discrete(limits = c("1", "0"),
                   #labels = c("BI", "BN"))

## Adding error bars
df_sum <- summarySE(df, measurevar="proportion_peerbody", groupvars=c("bi")) # creating a summary of the variables of interest to extract error bars using Rmisc package
df_sum

ggplot(df_sum, aes(x=bi, y=proportion_peerbody)) + # we use the summary we created to plot
    geom_bar(position=position_dodge(), stat="identity") +
    geom_errorbar(aes(ymin=proportion_peerbody-se,
                      ymax=proportion_peerbody+se), 
                  width=.2, position=position_dodge(.9))
```


## Examining interactions - bar graphs
This section walks through how to look at a interaction between dichotomous variables using a bar graph. After creating the base graph, you can make additional changes to asthetic elements such as changing colors, legends and bar direction.
```{r}
ggplot(df, aes(x=bi, y=proportion_socconv, fill = sex)) +
  geom_bar(stat = "identity", position = "dodge") 
  #scale_fill_brewer(palette = "Set2") # pre-designed color palettes
  #labs(fill = "Sex") 
  #scale_fill_manual(labels = c("Boys", "Girls"), # Rename fills
                    #values = c("#ff8c00", "#5898d7")) # select your own color options
  #theme(legend.justification=c(-0.1,1), legend.position=c(.75,.95)) 
  #coord_flip()
```


## Examining interactions - 3 variable interactions
This section provides an example of how to examine a three-way interaction where one variable is dichotomous.
```{r, warnings=FALSE}
ggplot(df, aes(x=proportion_toys, y=proportion_socconv, color = sex)) +
  geom_point() +
  stat_smooth(method = lm, se = FALSE) 
  #facet_wrap(~bi)
```


### Examining interactions - continuous
This section provides some advanced options for examining continuous interactions using the jtools package.
```{r}
# Run your model to feed into jtools functions
lm1 <- lm(proportion_socconv~biq*proportion_toys, df) 

# Creating a +/- 1SD plot
interact_plot(lm1, pred = biq, modx = proportion_toys,
              plot.points = TRUE,
              x.label = "Behavioral Inhibition",
              y.label = "Proportion of Time Engaging in Social Conversation",
              legend.main = "Prop Dwell Toys")

# Creating a regions of significance plot
johnson_neyman(lm1, pred = biq, modx = proportion_toys, alpha = 0.05)
```


## Publication quality graphs and exporting
This is an example of the work that goes into creating a plot worthy of publication. There are also a couple of ways to export your final plot depending on if you would like to make further edits in a program such as inkscape or illustrator. 
```{r, warnings=FALSE}
df %>%
  dplyr::select(id, bi, proportion_peerbody, proportion_socconv) %>%
  gather(key = type, value = proportion, proportion_peerbody, proportion_socconv) -> social

social_plot <- ggplot(social, aes(type, proportion, fill = bi)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_x_discrete(limits = c("proportion_peerbody", 
                              "proportion_socconv"), 
                   labels = c("Dwell Peer", "Social Conversation")) +
  labs(x = "Type of Behavior",
       y = "Proportion of Time Engaged in Behavior",
       fill = "Temperament") +
  scale_fill_manual(labels = c("BN", "BI"), values = c("#ff8c00", "#5898d7")) +
  theme_bw() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.spacing = unit(1.5, "lines")) +
  theme(axis.text = element_text(size = 12)) +
  theme(axis.title = element_text(size = 14)) +
  theme(legend.text = element_text(size = 12)) +
  theme(legend.title = element_text(size = 14)) +
  theme(legend.justification=c(-0.1,1), legend.position=c(.72,.95))

ggsave("../figures/social_plot.png", plot = social_plot, 
       width = 6, height = 6, dpi = 300) #make a 6 x 6 inch PNG file with 300 DPI

# Vector graphic: for editable figures in Inkscape or Illustrator
svg("../figures/social_plot.svg")
plot(social_plot)
dev.off()

#And for fun, just print it here (for html display)
plot(social_plot)
```

