---
title             : "R you ready to write a paper in R Markdown?"
shorttitle        : "R you ready"
author: 
  - name          : "Rick O. Gilmore"
    affiliation   : "1,2"
    corresponding : yes    # Define only one corresponding author
    address       : "Department of Psychology, The Pennsylvania State University, University Park, PA 16802 USA"
    email         : "rogilmore@psu.edu"
  - name          : "Michael Hallquist"
    affiliation   : "1"
affiliation:
  - id            : "1"
    institution   : "The Pennsylvania State University"
  - id            : "2"
    institution   : "Databrary.org"
author_note: |
  The authors are with the Department of Psychology at The Pennsylvania State University.
  The authors acknowledge support from the Department of Psychology, the Social, Life, & Engineering Sciences Imaging Center (SLEIC), and the Child Study Center's Open Data in Developmental Science (ODDS) initiative.
abstract: |
  Want to write a paper using R Markdown? Keep reading to see how.
  
keywords          : "APA, R Markdown"
wordcount         : "Not that many."
bibliography      : ["r-references.bib", "references.bib"]
figsintext        : no
figurelist        : no
tablelist         : no
footnotelist      : no
lineno            : yes
lang              : "english"
class             : "man"
output            : papaja::apa6_pdf
editor_options: 
  chunk_output_type: console
params:
  data_file_in: '../data/survey_clean.csv'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning=FALSE, error=FALSE)
```

```{r load_packages, include = FALSE}
if(!require("papaja")){devtools::install_github("crsh/papaja")}
library("papaja")

if(!require("afex")){install.packages("afex")}
library("afex")

if(!require("gmodels")){install.packages("gmodels")}
library("gmodels")

# May need this for newer versions of R (needed for R 3.4.1)
# devtools::install_github("crsh/papaja")

if(!require("tidyverse")){install.packages("tidyverse")}
library("tidyverse")

# Could also source("R/Install_load_package.R") and then
# lapply(c("papaja", "tidyverse"), Install_load_package)
```

```{r analysis_preferences}
# Seed for random number generation
set.seed(42)
```

It is possible to write an entire APA-formatted article in R Markdown. 
This very brief paper shows how it might be done.
As illustration, we use the data from a short, informal survey of participants in the 2018 R Bootcamp at Penn State.
<!-- We predicted that higher levels of enthusiasm for banjo music would be reported by respondents with *lower* reported hours/day of preferred sleep, at least among younger respondents. -->

```{r import-survey}
#survey <- read_csv("data/survey_clean.csv")
# Or we can use "cleaned" actual survey data
survey <- read_csv(params$data_file_in)

survey$R_exp <- ordered(survey$R_exp, levels=c("none",
                                               "limited",
                                               "some",
                                               "lots",
                                               "pro"))

survey$Sub_id <- 1:length(survey$R_exp)
```

# Methods {#methods}
Consistent with open and transparent science practices, we report how we determined our sample size, all data exclusions (if any), all manipulations, and all measures in the study [@Simmons2011-za].

## Participants {#participants}

We asked participants in an optional "R Bootcamp" held at the Pennsylvania State University Department of Psychology on August 16 and 17, 2018 to complete an anonymous survey using a Google Form.
We asked participants to report how old they felt.
A total of $n=$ `r length(survey$Psych_age_yrs)` respondents answered the survey with a reported felt age of $M$=`r mean(survey$Psych_age_yrs)` and a range of [`r min(survey$Psych_age_yrs)`-`r max(survey$Psych_age_yrs)`] years. 

## Material {#material}

The survey can be found at this URL: <https://docs.google.com/forms/d/e/1FAIpQLSeGqic9Hrj-XvkESZmu_0t6H02R-U6yzYnRLuX6HDFDp4R39g/viewform>.
There were six questions asked:

1. Your current level of experience/expertise with R
2. Your enthusiasm for banjo music?
3. How old do you feel (in years)?
4. Preferred number of hours spent sleeping/day
5. Favorite day of the week?
6. Is there a reproducibility 'crisis' in psychology?

## Procedure {#procedure}

We emailed a link to the survey to the list of participants in advance.
We also include a link to the survey on the web page containing the course schedule (https://psu-psychology.github.io/r-bootcamp-2018/schedule.html).
We encouraged participants to complete the survey before the first day or during lunch.

## Data analysis {#data-analysis}
We used `r cite_r("r-references.bib")` for all our analyses.
The code used to generate these analyses is embedded in this document.
To view it, see the R Markdown file in the [GitHub repository](http://github.com/psu-psychology/r-bootcamp-2018/talks) associated with this paper.

# Results {#results}

```{r Banjo-by-experience, results="asis"}
descriptives_Banjo <- survey %>%
  select(R_exp, Banjo) %>%
  group_by(R_exp) %>%
  summarise(Mean = mean(Banjo),
            Median = median(Banjo),
            SD = sd(Banjo),
            Min = min(Banjo),
            Max = max(Banjo))

# Drop the Timestamp column
descriptives_Banjo[, -1] <- printnum(descriptives_Banjo[, -1])

apa_table(
  descriptives_Banjo, 
  caption = "Descriptive statistics of banjo music enthusiasm by R experience.",
  note = "This table was created with apa_table()",
  escape = TRUE
)
```

```{r apaTables-corr, results="asis"}
# The apa_print commands produce LaTex, so we want those to be included "as-is"
corr_Banjo_age <- apa_print(cor.test(survey$Banjo,survey$Psych_age_yrs))
corr_Banjo_sleep <- apa_print(cor.test(survey$Banjo,survey$Sleep_hrs))
corr_age_sleep <- apa_print(cor.test(survey$Psych_age_yrs,survey$Sleep_hrs))
```

```{r apa-corr-table, results="asis"}
# Source custom function for correlation p-values
source('../R/corstars.R')

# Get correlation matrix
cor_mat <- round(survey[c('Banjo','Psych_age_yrs','Sleep_hrs')],2)

# Format with sig. stars using custom function
cor_stars <- as.data.frame(corstars(cor_mat))

apa_table(cor_stars,
          caption = "Correlation table of the example data set.",
          note = "This is a correlation table created using apa_table().",
          escape = TRUE)

```


```{r Banjo-by-age-exp, fig.cap="Banjo music enthusiasm by age and R experience"}
Banjo_vs_r_exp <- survey %>%
  ggplot() +
  aes(x=Banjo, y=Psych_age_yrs) +
  facet_grid(. ~ R_exp) +
  geom_point()
Banjo_vs_r_exp
```

```{r Banjo-by-sleep, fig.cap="Banjo music enthusiasm by preferred hours of sleep"}
Banjo_vs_sleep <- survey %>%
  ggplot() +
  aes(x=Sleep_hrs, y=Banjo) +
  geom_point() +
  geom_smooth() +
  theme_apa()
Banjo_vs_sleep
```

```{r Banjo-r_exp-aov}
Banjo_aov <- afex::aov_ez(id="Sub_id", dv="Banjo", data=survey, between = "R_exp")
Banjo_aov_results <- apa_print(Banjo_aov, es = "pes")
Banjo_aov_results_p <- apa_print(Banjo_aov, es = "pes", in_paren = TRUE)
```

```{r Banjo-aov-table, results="asis"}
# apa_table also returns LaTex, so render results "as-is"
apa_table(Banjo_aov_results$table,
          align = c("l", "r", "c", "r", "r", "r"),
          caption = "ANOVA table for the analyis of the example data set.",
          note = "This is a table created using apa_print() and apa_table().")
```

Table \@ref(tab:Banjo-by-experience) summarizes the banjo music enthusiasm ratings data by levels of R experience.
As Gilmore predicted, the more participants know about R, the more they come to appreciate banjo music. 

Let's examine the correlations between our continuous variables.
As indicated in Table \@ref(tab:apa-corr-table), there is a non-significant negative correlation (`r corr_Banjo_age$estimate`) between banjo music enthusiasm and age (`r corr_Banjo_age$statistic`), no correlation (`r corr_Banjo_sleep$estimate`) between banjo music enthusiasm and sleep (`r corr_Banjo_sleep$statistic`), but a positive correlation (`r corr_age_sleep$estimate`) between age and sleep (`r corr_age_sleep$statistic`). Figures \@ref(fig:Banjo-by-age-exp) and \@ref(fig:Banjo-by-sleep) depict these patterns.

To test the hypothesis that banjo music enthusiasm varies as a function of R expertise, we carried out a one-way ANOVA.
R experience (`r Banjo_aov_results$full$R_exp`) did not predict enthusiasm for banjo music, so Gilmore will have to continue searching for userRs who appreciate the banjo. 
Table \@ref(tab:Banjo-aov-table) summarizes these results.

# Discussion {#discussion}

These results aren't going to set the world on fire, but they do show how awesome it can be to use R, R Markdown, and literate programming principles to conduct and open, transparent, and reproducible psychological science. 
Yay, us!

There are no limitations to what we can accomplish using these tools. So, let's get to it.

<!-- This is a LaTex command -->
\newpage

# References

```{r create_r-references}
r_refs(file = "r-references.bib")
```

<!-- This is also LaTex code -->
\setlength{\parindent}{-0.5in}
\setlength{\leftskip}{0.5in}
