---
title: "Scaling Up with Functions"
author: "Daniel Albohn"
date: "`r Sys.time()`"
output:
  html_document:
    theme: spacelab
    toc: yes
    toc_depth: 3
    toc_float: TRUE
bibliography: bib/references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      tidy = TRUE,
                      collapse = FALSE,
                      warning = FALSE,
                      message = FALSE,
                      cache = FALSE)
```

# Establishing the Problem
    
    If you find yourself copy and pasting the same thing more than 3 times
    you should probably write a function to handle it.

Individuals who work in data science usually find themselves doing repetitive
tasks and analyses. How can we: 1) Automate to prevent copy-pasting? and
2) Scale up something (e.g., script, function) that works on a lower level to 
work at a higher level (e.g, on multiple files or input).

Some examples of common problems that might be answered by "scaling up":

  - Reading multiple data files into your active Environment
  
  - Fitting multiple models to multiples observations
  
  - Obtaining multiple graphs for each subject
  
  - Applying a custom function

# Session Goals
The goals of this session are two-fold:

  1) Gain a deeper understanding of functions,
  
  2) Learn how to apply functions to "Scale it Up"

## Before We Begin
Make sure you have the following packages installed if you would like to follow along.

```{r load}
# Load and/or install
pacman::p_load(tidyverse, broom, gapminder)
```

# Functions
Functions consist of a `formal` and a `body`, and take the form
`my_function(formal){body}`. You must always assign a
function to an object to be able to use it, unless you use an
anonymous function.

```{r}
hello_world <- function() { print('Hello, world!') }
hello_world()
```

Congratulations! Now let's take it up a notch...

```{r}
## When on multiple lines, functions run downstream...
hello_world_name <- function(name){
  message <- paste0('Hello, ',name,'!') # paste0 is paste without any spaces
  return(message)
}

hello_world_name('Dan')
```

A note about `return()`. If you do not manually specify what should be
returned from a function, R will assume that it is the result of the last
line in the function.

If you want to return something from a function that is more than a single
element (e.g., a data frame), you will need to assign it to a list, for
example.

# Begin to Scale it Up: `apply` and `map`
Apply functions are essentially `for` loops simplified. These functions
take a list as input and supply a list as output.

```{r}
list <- lapply(1:10, print)
```

```{r}
lapply(list, sqrt)
```

Map functions from the `purrr` package are like apply, but have a few
advantages. For example, you don't have to always have to write out
functions, and there are type specific `map` functions.

```{r}
my_list <- list(letters=letters, numbers=1:26)
```

So, maybe we want to extract the 3rd element from each part of the list...
This won't work:

```{r eval=FALSE}
lapply(my_list, 3)
```

    Error in match.fun(FUN) : '3' is not a function, character or symbol

But this does!

```{r}
map(my_list, 3)
```

And this!

```{r}
my_list %>% 
  map(., 3)
```

If you wanted to stay in base R, you'd have to do this:

```{r}
lapply(my_list, '[[', 3)
```

The `[[` should look familiar as the left half of the `[[]]` from the indexing tutorial.

## More `map`: Variations

```{r, eval=FALSE}
?map
```

Base `map` takes a list as input and returns a list:

```{r}
map(1:10, print)
```

`map_df` takes a df as input, returns a df

```{r}
df <- data.frame(x=1:10,y=11:20,z=letters[1:10])
map_df(df, mean)
```

`map_if` applies a function if a condition is met, and if not will return the object
unaltered.

```{r}
map_if(df, is.integer, mean)
```

We can pass additional arguments like so:

```{r}
?scale # Takes three arguments

map_df(df[1:2], scale, center=FALSE)

# Alternatively
# map(df[1:2], scale, center=FALSE)
```

## More `map`: `map2`
`map2` allows for two inputs to be present for a function.

```{r}
map2(1:10, 2, function(x,y) {x * y})
```

## Anonymous functions
One of the benefits of `map` over `*apply` is that we can use short hand,
or anonymous functions. If we have to explicitly define a function in
`map` if takes this form:

```{r}
map2(1:10,2,function(x,y) {x * y})
```

But, we can take a short cut and write an anonymous function in this
form:

```{r}
map2(1:10, 2, ~.x * .y)
```

## A Quick Scaling Up Example
All of his mapping is very basic and it might be hard to understand the power of
mapping functions to lists. As a quick example, let's examine a more practical
solution to a common problem from one of the issues brought up at the beginning
of this tutorial.

I have a folder of data (looking at you `EPrime`!) and I want to read all of the contents
into `R`.

```{r scale-up1}
source('../R/scale_up_utils.R')

# Function to create lots of data in a folder
my_dir <- lots_of_data()

# List the files in the folder
file_names <- list.files(my_dir, pattern = "*.csv", full.names = TRUE)

# Read it in
df <- purrr::map(file_names, readr::read_csv)
length(df)
```

What is this black magic? Well, like we saw above, we now have a special
list with length = 30 and each element is a separate data frame. We can see each
individual data frame by subsetting our list.

```{r scale-up2}
head(df[[1]])
```

This might be useful for some analyses, say, for example if you want to run analyses
on the participant level, but more often than not we want a 2-dimensional data frame that
is in tidy data format. To do so, we need to reduce or simplify our nested list of many
data frames. To do so, we use the verb `reduce`.

`reduce` simplifies list elements by combining common elements via a user-specified
function. In this case we will use `rbind` so that our data is combined by stacking
each data frame on top of each other.

```{r}
df_reduced <- df %>% 
  reduce(., rbind)

df_reduced[c(1:4,300:304),]
```

## Getting data from a folder (start to finish)
How succinct can we make this process? Fairly short if we assume we already have the
raw data in a folder.

```{r scale-up-all, eval=FALSE}
# The whole thing!
df <- list.files(my_dir, pattern = "*.csv", full.names = TRUE) %>% 
  map(., read_csv) %>% 
  reduce(., rbind)
```

Utilizing pipes and `tidyverse` verbs, we can accomplish a fairly complex task in three lines
of code.

Make sure to delete the random folder we created so it isn't taking up unnecessary space.

```{r clean-up, eval=FALSE}
# Clean up
unlink(my_dir, recursive = TRUE)
```

# Our Gapminder Scaling Problem
Let's try scaling something else on an issue that might be a bit more complicated. Here,
we will use the `gapminder` data set to fit individual linear models to many different
groups of observations.

A large majority of this section of the tutorial is based on the
chapter [`Many Models`](http://r4ds.had.co.nz/many-models.html)
from Hadley Wickham's wonderful book [`R for Data Science`](http://r4ds.had.co.nz/)[@wickham2017r].

## The Gapminder data set
This is a data set that lives in a package based on world economic, health, etc. data
collected by the ["Gapminder"](http://www.gapminder.org/data/) Foundation. Gapminder
is a non-profit Swedish organization aimed at educating consumers with evidenced-based
facts about the world.

The Gapminder website has many, many [data files](https://www.gapminder.org/data/)
available for public use, and lots of 
[cool visualizations](https://www.gapminder.org/tools/#$chart-type=bubbles)
to play around with. However, we will be most interested in the following:

```{r gapminder, echo=FALSE}
library(kableExtra)
classes <- map_df(gapminder::gapminder, class) %>% gather()

gap <- data.frame(Variable = names(gapminder::gapminder),
                  Class = classes$value,
                  Description = c("Country", "Continent of Country", "Year data was sampled",
                                  "Life expectancy at birth in years", "Population",
                                  "GDP per capita (adjusted)")
                  )
knitr::kable(gap, format = 'html') %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed", "responsive"),
                full_width = TRUE,
                font_size = 15)
  
```

## The `broom` package
This package from David Robinson provides model output from a variety of
model functions in a tidy format. We will be using `broom` as we develop a
pipeline and scale up our functions.

```{r cite}
# You can check citations in R if you can't remember!
citation('broom')
```

## Mind the "gap" 
Because this data frame has many types of data (numeric, time, groups), we can begin to ask and look
at very cool things ("PLOT YOUR DATA!").

**Question**: Is GDP increasing across all countries?

Start by plotting the data:

```{r problem1}
gapminder %>% 
  ggplot(., aes(year, gdpPercap, group = country)) + 
  geom_line(alpha = 0.3, size = 2) + 
  theme_bw(base_size = 15)
```

Oh, my. something is happening with a few countries. I wonder what continent the most
 extreme trends are on:

```{r problem2}
gapminder %>% 
  ggplot(., aes(year, gdpPercap, group = country, colour = continent)) + 
  geom_line(alpha = 0.3, size = 2) + 
  theme_bw(base_size = 15)
```

Let's pull out one country and see if we can find the abnormal culprit.

```{r problem3}
gapminder %>% 
  filter(gdpPercap >= 90000) %>% 
  distinct(country) # We only need to see it once
```

Well, well, well...ok, let's plot just this country and begin to fit just one model.

```{r problem4, fig.width=10, echo=FALSE}
kw <- filter(gapminder, country == "Kuwait")

p1 <- kw %>% 
  ggplot(aes(year, gdpPercap)) + 
  geom_line(size = 2) + 
  ggtitle("Full data        = ") +
  theme_bw(base_size = 15)

kw_mod <- lm(gdpPercap ~ year, data = kw)
p2 <- kw %>% 
  modelr::add_predictions(kw_mod) %>%
  ggplot(aes(year, pred)) + 
  geom_line(aes(kw$year, kw$gdpPercap), alpha = .5, colour='red', size = 2) +
  geom_line(size = 2) + 
  ggtitle("Linear trend      + ") +
  theme_bw(base_size = 15)

p3 <- kw %>% 
  modelr::add_residuals(kw_mod) %>% 
  ggplot(aes(year, resid)) + 
  geom_hline(yintercept = 0, colour = "white", size = 2) + 
  geom_line(size = 2) + 
  ggtitle("Remaining pattern") +
  theme_bw(base_size = 15)

cowplot::plot_grid(p1,p2,p3, align = 'h', nrow = 1)
```

Let's compare Kuwait to another country (Canada):

```{r problem5, fig.width=10, echo = FALSE}
ca <- filter(gapminder, country == "Canada")

p4 <- ca %>% 
  ggplot(aes(year, gdpPercap)) + 
  geom_line(size = 2) + 
  ggtitle("Full data        = ") +
  theme_bw(base_size = 15)

ca_mod <- lm(gdpPercap ~ year, data =ca)
p5 <- ca %>% 
  modelr::add_predictions(ca_mod) %>%
  ggplot(aes(year, pred)) + 
  geom_line(aes(ca$year, ca$gdpPercap), alpha = .5, colour = 'red', size = 2) +
  geom_line(size = 2) + 
  ggtitle("Linear trend      + ") +
  theme_bw(base_size = 15)

p6 <- ca %>% 
  modelr::add_residuals(ca_mod) %>% 
  ggplot(aes(year, resid)) + 
  geom_hline(yintercept = 0, colour = "white", size = 3) + 
  geom_line(size = 2) + 
  ggtitle("Remaining pattern") +
  theme_bw(base_size = 15)

cowplot::plot_grid(p4,p5,p6, align = 'h', nrow = 1)
```

It looks like a linear trend fits Canada's GRP over the years quite well!

## Starting simple: A single solution
OK, so how do we fit that linear model to just one country (e.g., Kuwait or Canada)? Well, first we
need to subset or filter our data by country.

```{r}
kw <- filter(gapminder, country == 'Kuwait')
head(kw)
```

Now we can apply our linear model to just Kuwait's data.

```{r}
fit1 <- lm(gdpPercap~year, data = kw)
fit_summary <- summary(fit1); print(fit1)
```

And we can clean this up a bit with `broom`'s `tidy` function.

```{r}
clean_summary <- tidy(fit_summary); print(clean_summary)
```

We can also "augment" our original data frame with some useful additions with
`broom`'s `augment` function.

```{r}
fit_augment <- fit1 %>% augment(.)
head(fit_augment)
```

```{r}
fit_augment %>% 
  ggplot(., aes(year, .fitted)) + 
  geom_line(size = 2) + 
  geom_line(aes(fit_augment$year, fit_augment$gdpPercap), size = 2, colour = 'red')
```

This should look familiar from above. So, it looks like we successfully fitted a model to Kuwait's
data. But what if we want/need to compute a separate linear model for each country? Well, we have
a few options:

  1). Copy and paste the above code for each country
  
  2). Develop some sort of `for` loop or `apply` script with a custom linear model function
  
Well, the first option seems intuitive and logical, but how many countries are there? That will be
how many times we have to copy and paste our above code.

```{r}
gapminder %>% 
  select(., country) %>% 
  distinct(., country) %>% 
  tally()
```

Hm, copy-pasting 142 times seems ridiculous and not a logical solution. How about a `for` loop? I will be
using `broom`'s `glance` function to provide a single line for each model some summary statistics
that will be useful for comparing model fit.

```{r}
countries <- gapminder %>% 
  select(., country) %>% 
  distinct(., country)

tidy_fits <- data.frame()

for (x in 1:nrow(countries)) {
  countryx <- countries[x,]
  df <- filter(gapminder, country == countryx$country)
  
  fitx <- lm(gdpPercap~year, data = df)
  xtidy <- glance(fitx)
  xtidy$country <- countryx$country
  
  tidy_fits <- rbind(tidy_fits, xtidy)
}

tidy_fits %>% select(., country, everything()) %>% head(.)
```

OK...that's acceptable, and it can help us conceptually understand what we need to do in order to
accomplish our task, but it's clunky and intuitive. Isn't there anything else?

## The `tidyverse` Scale Up
Well, we do have a third option:

  3). Some `tidyverse` alternative which is probably easier (spoiler: it is)

### Nested data frames
You might imagine that using `map` might help us both consolidate our code and make it more
readable for future reference. Conceptually, we want to do the exact same as the `for`
loop above: extract grouped data by country, apply a linear model to each country's data,
and then add the tidied linear model summary to a tidy data frame.

In order to do this within a `tidyverse` framework, we need to *nest* our data frame
by our grouping variable. A nested data frame is a special data frame that has for
each observation (row) another data frame with only the observations of split by
the grouping variable.

```{r}
by_country <- gapminder %>% 
  group_by(., country) %>% 
  nest(.)

head(by_country[[1,2]])
```

Next, we need to do some setup and create a specialized linear model function

```{r}
country_model <- function(x) {
  lm(gdpPercap~year, data = x)
}
```

That's it! Just wrapping our linear model in a nice `function`! Now we are ready to apply it to each of
our nested data frames.

```{r}
models <- map(by_country$data, country_model)
```

We can now add to this nested data frame a model column that will contain all of the
model output for each country.

```{r}
by_country <- by_country %>% 
  mutate(model = map(data, country_model))

head(by_country)
```

Finally, we can get our comparison list by applying `glace` to each country's model and then
*unnesting* the data.

```{r}
tidy_fits2 <- by_country %>% 
  mutate(glance = map(model, broom::glance)) %>% 
  unnest(glance)

head(tidy_fits2)
```

# References