---
title: "Basics of ANOVA and categorical data analysis in R"
author: "Michael Hallquist"
date: "15 Aug 2018"
output:
  html_document:
    code_folding: show
    df_print: kable
    mathjax: default
    number_sections: yes
    theme: spacelab
    toc: yes
  pdf_document:
    code_folding: show
    df_print: kable
    number_sections: yes
    toc: yes
    toc_depth: 4
---
<style type="text/css">
body{ font-size: 20px; max-width: 1700px; margin: auto; padding: 1em; }
code.r{ font-size: 22px; }
pre { font-size: 18px; }
</style>

```{r setup, include=FALSE}
if (!require(pacman)) { install.packages("pacman"); library(pacman) }
p_load(knitr, tidyverse, ggfortify, car, Hmisc, stargazer, apaTables, jtools, emmeans, ez, afex, multcomp,
       vcd, vcdExtra, GGally, gmodels, xtable)

knitr::opts_chunk$set(echo = TRUE) #print code by default
options(digits=3)
data(sk2011.1)
```

# Introduction

The overall goal is to review ANOVA methods in R, as well as analyses of contingency tables (categorical data).

# ANOVA

## Simple between-subjects designs
For between-subjects designs, the `aov` function in R gives you most of what you'd need to compute standard ANOVA statistics. But it requires a fairly detailed understanding of sum of squares and typically assumes a balanced design. The `car::Anova` function takes things a bit further by allowing you to specify Type II or III sum of squares.

Consider the Moore 1971 dataset on conformity responses as a function of partner status (high/low) and level on an authoritarianism scale (highm medium, low). This is a between subjects design with 2 factors (2 x 3).

```{r}
mod <- lm(conformity ~ fcategory*partner.status, data=Moore)
Anova(mod, type="III")
```

Even better is the `afex` package, which handles most of the complexity of ANOVA designs.

```{r}
Moore$id <- 1:nrow(Moore)
res <- aov_ez(id="id", dv = "conformity", between = c("partner.status", "fcategory"), data=Moore)
print(res)
```

More detailed coverage of this example: <https://cran.r-project.org/web/packages/afex/vignettes/afex_anova_example.html>

## User-friendly coverage of all ANOVA-type designs

In general, the `aov_ez` function from the `afex` package is an ideal tool for ANOVA analysis because it computes the expected ANOVA table, as well as effect size (generalized eta squared).

Let's consider the experiment of Singmann and Klauer (2011), where they examined the conditional reasoning of individuals based on instruction type (b/w subs: deductive versus probailistic), inference validity (w/i subs: valid versus invalid problems) and plausibility (plausible versus implausible). In short, this is a 2 x 2 x 2 mixed ANOVA design that can be tested without much difficulty.

```{r}
str(sk2011.1)
```

```{r}
a1 <- aov_ez("id", "response", sk2011.1, between = "instruction", 
       within = c("inference", "plausibility"))

summary(a1)
```

Pretty print out of ANOVA results

```{r}
knitr::kable(nice(a1))
```

What about the estimated marginal means of the design?

```{r}
emmeans(a1, ~instruction * inference * plausibility)
```

What about the 'main effects?'

```{r}
em1 <- emmeans(a1, ~inference)
```

And pairwise difference among levels of inference validity? 

```{r}
pairs(em1)
```

Note that emmeans plays well with the multcomp package, which provides a broader suite of multiple comparison corrections:

```{r}
summary(as.glht(pairs(em1)), test=adjusted("free"))
```

## Plotting results of aov_ez

```{r}
emmip(a1, instruction ~ inference|plausibility, CIs = TRUE)
```

The `emmeans` package provides the `emmip` function, which is very useful for plotting the results of an `aov_ez` object.

# Working with categorical data

The `vcd` and `vcdExtra packages are particulary handy for working with categorical data.

```{r}
data("Arthritis")

# attach(Arthritis)
# ###
# Treatment
# detach(Arthritis)

#table(Arthritis$Treatment, Arthritis$Sex)
with(Arthritis, table(Treatment, Sex))

tb <- with(Arthritis, table(Treatment, Sex))

tb["Placebo", "Female"]

with(Arthritis, table(Treatment))

#cell proportions
round(prop.table(tb), 3)

#row proportions
round(prop.table(tb, 1), 3)

#column proportions
round(prop.table(tb, 2), 3)

#sum over rows
margin.table(tb, 1)

#add row margins
addmargins(tb, 1)
addmargins(tb, 2)
addmargins(tb, c(1,2))

#3-way table with formula syntax
#tb3 <- xtabs(~Treatment + Improved + Sex, Arthritis)
tb3 <- xtabs(~Treatment + Sex + Improved, Arthritis)
tb3
ftable(tb3)

## col ~ row
structable(Sex + Treatment ~  Improved, Arthritis)

xtable(xtabs(~ Sex + Treatment, Arthritis))

#sum frequencies across sexes
freqdf <- Arthritis %>% count(Improved, Treatment) %>% ungroup()

#proportions
Arthritis %>% count(Improved, Treatment) %>% mutate(prop = prop.table(n))

#can obtain frequency tables from dplyr, too!
Arthritis %>%
  count(Improved, Treatment) %>%
  spread(Treatment, n)

Arthritis %>%
  count(Improved, Treatment, Sex) %>%
  spread(Treatment, n)

#same as
xtabs(~Improved + Treatment, Arthritis)

#and can collapse over factors in frequency form
freqdf %>% count(Improved, wt=n) #tell it to weight the counts by frequency variable n
freqdf %>% group_by(Improved) %>% dplyr::summarize(freq=sum(n)) #more explicit syntax

freqdf %>% count(Treatment, wt=n)

#2-way chi-square test
chisq.test(xtabs(~Treatment + Improved, Arthritis))

#association statistics
assocstats(xtabs(~Treatment + Improved, Arthritis))

#mosaic(structable(Treatment+Improved ~ Sex, Arthritis))
mosaic(structable(Sex+Treatment~Improved, Arthritis))

ggpairs(dplyr::select(Arthritis, -ID))


tv.data<-read.table(system.file("doc","extdata","tv.dat",package="vcdExtra"))
TV <- xtabs(V5 ~ ., data=tv.data)

dimnames(TV) <- list(Day=c("Monday","Tuesday","Wednesday","Thursday","Friday"),
                     Time=c("8:00","8:15","8:30","8:45","9:00","9:15","9:30",
                            "9:45","10:00","10:15","10:30"),
                     Network=c("ABC","CBS","NBC","Fox","Other"),
                     State=c("Off","Switch","Persist"))

ggpairs(TV)


TV <- TV[,,c("ABC", "CBS", "NBC"),]     # keep only ABC, CBS, NBC
TV <- TV[,,,"Persist"] #people who keep watching
structable(TV)

ggpairs(TV)

```

```{r, size="tiny"}
#more SPSS-like
CrossTable(TV[,,1],prop.t=FALSE,prop.r=FALSE,prop.c=FALSE)

```

## Logistic regression

We won't cover this here, but check out this nice tutorial: <https://stats.idre.ucla.edu/r/dae/logit-regression/>

## Multinomial regression

Likewise, see this: <https://stats.idre.ucla.edu/r/dae/multinomial-logistic-regression/>