---
title: "ggplot2 tutorial - R Bootcamp 2018"
author: "Alicia Vallorani"
date: "August 13, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

if (!("pacman" %in% installed.packages()[,"Package"])) { install.packages("pacman") } #handy package manager that installs and loads packages used in this document
pacman::p_load(dplyr, tidyr, ggplot2, Hmisc, GGally, jtools) #install (if needed) and load packages
```

## Read in data
```{r}
df <- read.csv("../data/ggplot2_tutorial_vallorani.csv", stringsAsFactors = FALSE) %>%
  mutate_at(vars(bi, sex, socbid_group, prosoc_group), as.factor) ### Consider clarifying the levels of the group and bi variables so the data.frame is self-describing
```

## Examine data structure
```{r}
str(df)
```

## Examining univariate distributions - histograms
```{r}
# Looking at a histogram for a single variable
ggplot(df, aes(proportion_toys)) +
  geom_histogram(bins = 5) # you can change the bin value to best fit your data

# Looking at histograms for all variables ggplot option
### Alicia, I wasn't sure which columns should be included since binary don't make sense
###      (this was throwing an error). I Took a guess here
ggplot(df %>% select(starts_with("proportion"), age) %>% gather(), aes(value)) +
    geom_histogram(bins = 5) + 
    facet_wrap(~key, scales = "free_x")

```

### Alternative approach to histograms for all variables using Hmisc package
```{r, fig.height=9, fig.width=8}
hist.data.frame(df %>% select(-id, -did)) ### I'm ambivalent about the value of showing people 2 ways of doing the same thing in an introductory walkthrough
```

## Examining zero-order relations - scatterplots
```{r, fig.height=5, fig.width=10, messages=FALSE, warnings=FALSE}
# Looking at a scatterplot for a single relation
ggplot(df, aes(proportion_toys, proportion_socconv)) +
  geom_point()
  #stat_smooth() 
  #stat_smooth(method = "lm", color = "black", se = TRUE)
  #facet_wrap(~bi)
  #facet_wrap(~socbid_group)
  #face_grid(bi~sex)
```

```{r, fig.height=10, fig.width=10, messages=FALSE, warnings=FALSE}
# Looking at a scatterplot matrix
# Michael, I've never really found a ggplot option I am hapy with. Do you have good code?
### Do you mean customizing the overall appearance of ggpairs?

# Fancy distribution/relations option with GGally
ggpairs(df %>% select(-id, -did) %>% na.omit(), progress=FALSE, lower = list(combo = wrap("facethist", bins=6))) ### Consider plotting a somewhat smaller variable set for aiding discussion (e.g., 6 x 6)
```


## Examining group differences - bar graphs
```{r}
ggplot(df, aes(x=bi, y=proportion_peerbody)) +
  geom_bar(stat = "identity") 
  #theme_bw() 
  #theme(panel.grid.major.x = element_blank(),
        #panel.grid.minor.x = element_blank(),
        #panel.grid.major.y = element_blank(),
        #panel.grid.minor.y = element_blank(),
        #panel.spacing = unit(1.5, "lines")) 
  #labs(x = "Behavioral Inhibition Group",
       #y = "Proportion of Time Looking at Peer Body") 
  #scale_x_discrete(limits = c("1", "0"),
                   #labels = c("BI", "BN"))
```


## Examining interactions - bar graphs
```{r}
ggplot(df, aes(x=bi, y=proportion_socconv, fill = sex)) +
  geom_bar(stat = "identity", position = "dodge") #+
  #labs(fill = "Sex") +
  #scale_fill_manual(labels = c("Boys", "Girls"), values = c("#ff8c00", "#5898d7")) + ### What do you think about highlighting some of the built-in palettes (e.g., in the scale_fill_brewer function)?
  #theme(legend.justification=c(-0.1,1), legend.position=c(.75,.95)) +
  #coord_flip()

### I guess this dataset may not lend itself to this, but showing error bars (geom_errorbar) is a common request.
```


## Examining interactions - 3 variables and continuous interactions
```{r, warnings=FALSE}
ggplot(df, aes(x=proportion_toys, y=proportion_socconv, color = sex)) +
  geom_point() +
  stat_smooth(method = lm, se = FALSE) 
  #facet_wrap(~bi)
```

### Add subheading detail here
```{r}
# Advanced options using jtools
lm1 <- lm(proportion_socconv~biq*proportion_toys, df)

interact_plot(lm1, pred = biq, modx = proportion_toys,
              plot.points = TRUE,
              x.label = "Behavioral Inhibition",
              y.label = "Proportion of Time Engaging in Social Conversation",
              legend.main = "Prop Dwell Toys")

johnson_neyman(lm1, pred = biq, modx = proportion_toys, alpha = 0.05)
```


## Publication quality graphs and exporting
```{r, warnings=FALSE}
### you'll probably need to document/describe right-hand assignment, since it's a bit unconventional
df %>%
  dplyr::select(id, bi, proportion_peerbody, proportion_socconv) %>%
  gather(key = type, value = proportion, proportion_peerbody, proportion_socconv) -> social

social_plot <- ggplot(social, aes(type, proportion, fill = bi)) +
  geom_bar(stat = "identity", position = position_dodge()) +
  scale_x_discrete(limits = c("proportion_peerbody", 
                              "proportion_socconv"), 
                   labels = c("Dwell Peer", "Social Conversation")) +
  labs(x = "Type of Behavior",
       y = "Proportion of Time Engaged in Behavior",
       fill = "Temperament") +
  scale_fill_manual(labels = c("BN", "BI"), values = c("#ff8c00", "#5898d7")) +
  theme_bw() +
  theme(panel.grid.major.x = element_blank(),
        panel.grid.minor.x = element_blank(),
        panel.grid.major.y = element_blank(),
        panel.grid.minor.y = element_blank(),
        panel.spacing = unit(1.5, "lines")) +
  theme(axis.text = element_text(size = 12)) +
  theme(axis.title = element_text(size = 14)) +
  theme(legend.text = element_text(size = 12)) +
  theme(legend.title = element_text(size = 14)) +
  theme(legend.justification=c(-0.1,1), legend.position=c(.72,.95))

ggsave("../figures/social_plot.png", plot = social_plot, 
       width = 6, height = 6, dpi = 300) #make a 6 x 6 inch PNG file with 300 DPI

# Vector graphic: for editable figures in Inkscape or Illustrator
svg("../figures/social_plot.svg")
plot(social_plot)
dev.off()
```

